Solving an Exam Scheduling Problem Using a Genetic Algorithm

### 2009-12-02 Paper/paper.tex
Paper update

\begin{abstract}
  We examine the exam scheduling problem with soft constraints for problems of varying
  sizes using a genetic algorithm, random search, hillclimbing, and a variant of simulated
  annealing. This is done in the context of maximizing the fitness of a schedule, where the
  fitness is relatively computationally expensive to compute for any schedule. The genetic
  algorithm is discussed in detail. We consider selecting parameters for the GA that
  maximize the quality of the schedule found after a constant number of evaluations. The
  GA, with a naive crossover operator, performs much better than the random search, but
  simulated annealing is far superior in this search space.
\end{abstract}

@@

  We define two quality functions, mapping schedules to real numbers in [0,1],
  one for students and one for rooms. These are meant to capture how much they
  "like" the schedule under consideration. (For instance, a student will rate
  poorly any schedule where she has 2 consecutive exams, and very poorly any 
  schedule which expects her to take two exams at the same time.)

@@

      If a different representation were chosen, such that the mutated schedules were
      different from, but likely to have quality similar to their parents, the GA would 
      likely be able to perform much better than it does, but we could not discover such a
      representation and are satisfied with this method for present purposes.

@@

    \subsection{Scheduling Problem Instances}
      We generated 2 sets of 3 scheduling problem instances to test the algorithms above with. 
      The package \texttt{Schedule Instance Generator} in our code repository manages the generation of 
      random problem instances with many free parameters to specify characteristics of the 
      output instance.

      The first set of instances were generated with the parameters in the table below. 
      The rooms and courses and students were given arbitrary names and the courses that each 
      student attended were selected with uniform probability from the complete set of courses.

      The second set was generated with a sort of correlation between the courses that each student 
      took, rendering the dataset more true to the sort of data that would be found in real-world 
      instances of the exam scheduling problem. The set of courses that each student takes was 
      selected in the following way. Each student was assigned, with uniform probability, a \emph{major} 
      and a random number of courses between 2 and 6. Each course was then selected, with probability 
      0.7, from among the $|S|/\#Majors$ courses in their major, or, with probability 0.3, uniformly from 
      the set of all courses. We can expect that each student takes 70\% of their courses from a small 
      subset of the courses, representing the field that they specialize in, and 30\% from other fields. 
      This simulates the way in which we expect student bodies to choose their courses, though the true 
      data is likely to exhibit far more structure, due to program requirements, notorious professors, etc.

      The 3 sizes vary from manageable to large. The smallest could potentially be solved exactly, 
      with methods not under consideration here, while the others have search spaces much too large 
      for any hope of an exact solution. For the largest, which simulates a UofT exam schedule, a 
      single evaluation of the fitness function took 0.2 seconds on the fastest machine owned by any 
      of the authors.

      The following table characterized the problem instances that we used 
      (with \textsc{SSS} being an approximation of the search space size):      
      \definecolor{light-gray}{gray}{0.87}
      \rowcolors{1}{light-gray}{white}
      \begin{center}
      \begin{small}
      \begin{tabular}{*{7}{c}}
      \hline 
      	& S 	& M 	& L 	& $S_{Cor}$ & $M_{Cor}$ & $L_{Cor}$ \\ 
      \hline \hline
       d 		& 7 	& 10	& 7 	& 7 	& 10 	& 7 	\\ 
       t 		& 5		& 8 	& 8 	& 5 	& 8 	& 8 	\\ 
       $|C|$	& 20 	& 200 	& 603 	& 20 	& 200 	& 600 	\\ 
       $|R|$	& 4 	& 10 	& 43 	& 4 	& 10 	& 43	\\ 
       $|S|$	& 50 	& 300 	& 21945 & 50 	& 300 	& 21945	\\ 
       $|L_s|$	& 1-5 	& 1-6 	& 1-5 	& 2-6 	& 2-6 	& 2-6 	\\ 
       \begin{tiny}$\#$Maj\end{tiny}	& N/A 	& N/A 	& N/A 	& 4 	& 10 	& 20 	\\ 
       \textsc{SSS}	& $10^{182}$ & $10^{1840}$ & $10^{6695}$ & $10^{182}$ & $10^{1840}$ & $10^{6689}$ \\
      \hline 
	  \end{tabular} 
	  \end{small}
	  \end{center}

@@

        In the graph below, each data point represents the average of 5 trials. To make 
        the scale clearer, the colour on the surface corresponds to the fitness, where 
        the highest fitnesses are red. The number of copies is fixed at 1, and the proportions 
        of mutations and crossovers are shown on the horizontal axes.

        \begin{figure}[H]
  		  \centering
  		  \includegraphics[width=8cm]{GAParametersData}
  	      \caption{Effectiveness of the GA's proportion parameters}
  	      \label{graph:GAParam1}
		\end{figure}

        The graph above indicates that for a fixed number of evaluations, best results are 
        achieved by selecting parameters that have the majority of individuals in the next 
        generation be generated by crossovers of the most fit schedules in the previous 
        generation. In order to achieve the very highest fitness, a few randoms or mutations 
        should also be incorporated.

@@

        %\begin{figure}[H]
  		%  \centering
  		%  \includegraphics[width=8cm]{MEDIUM-GA_Size_Data}
  	    %  \caption{Effectiveness of the GA's population parameters}
  	    %  \label{graph:GAParam2}
		%\end{figure}

  \subsection{Comparison Between the Different Searches}
    We tested the 4 search methods on all 6 of our peoblem instances. The results for 
    only two of the problems are shown, she small and large problems, since the medium 
    problem exhibited characteristics very similar to the large problem.

    Since all methods produced nearly indistinguishable results for the uncorrelated 
    versions. This suggests that it may be possible to use the expected internal 
    correlations more effectively that the methods we have proposed do.

@@

    \begin{figure}[H]
  	  \centering
  	  \includegraphics[width=8cm]{SMALL-TYPE2_FullAverageStats}
      \caption{Comparison of the different searches on a small data set}
	\end{figure}

	\begin{figure}[H]
  	  \centering
  	  \includegraphics[width=8cm]{HUGE-TYPE2_FullAverageStats}
      \caption{Comparison of the different searches on a large data set}
	\end{figure}

@@

\section{Conclusion}
  [\ldots]

>>>>

\begin{abstract}
  We examine the exam scheduling problem with soft constraints for problems of varying
  sizes using a genetic algorithm (GA), random search, hillclimbing, and a variant of simulated
  annealing. This is done in the context of maximizing the fitness of a schedule, where the
  fitness is relatively computationally expensive to compute for any schedule. The genetic
  algorithm is discussed in detail. We consider selecting parameters for the GA that
  maximize the quality of the schedule found after a constant number of fitness computations. 
  The GA, with a naive crossover operator, performs much better than the random search, but
  simulated annealing is far superior in this search space.
\end{abstract}

@@

  We define two quality functions, mapping schedules to real numbers in [0,1],
  one for students and one for rooms. These are meant to capture how much they
  "like" the schedule under consideration. (For instance, a student will rate
  poorly any schedule where she has 2 consecutive exams, and very poorly any 
  schedule which expects her to take two exams at the same time. A room will 
  rate poorly a schedule that assigns two different exams to be held in it at 
  the same moment).

@@

      If a different representation were chosen, such that the children schedules were
      different from, but likely to have quality similar to their parents, the GA would 
      likely be able to perform much better than it does, but we could not discover such a
      representation and are satisfied with this method for present purposes.

@@

    \subsection{Scheduling Problem Instances}
      We generated 2 sets of 3 scheduling problem instances with which to test the 
      above algorithms. The first set of instances were generated with the parameters 
      in the first three columns of the table below. The rooms, courses and students 
      were given arbitrary names and the courses that each student attended were 
      selected with uniform probability from the complete set of courses.

      The second set was generated with a sort of correlation between the courses that each 
      student took, rendering the dataset more true to the sort of data that would be found 
      in real-world situations. The set of courses that each student takes was selected such 
      that we could expect that 70\% of their courses were courses in their major. This 
      simulates the way in which we expect student bodies to choose their courses, though 
      the true data is likely to exhibit far more structure, due to program requirements, 
      notorious professors, etc.

      The 3 instances' size vary from manageable to large. The smallest could potentially be 
      solved exactly, with methods not under consideration here, while the others have 
      search spaces much too large for any hope of finding a perfect solution. For the 
      largest, which simulates a UofT exam schedule, a single evaluation of the fitness 
      function takes 0.2 seconds on on a typical desktop computer. This makes it difficult 
      to thoroughly investigate this instance.

      The following table characterizes the problem instances that we used ($\#$Maj is the 
      number of different majors used and \textsc{SSS} is an approximation of the search space size):    
      \definecolor{light-gray}{gray}{0.87}
      \rowcolors{1}{light-gray}{white}
      \begin{center}
      \begin{small}
      \begin{tabular}{*{7}{c}}
      \hline 
      	& S 	& M 	& L 	& $S_{Cor}$ & $M_{Cor}$ & $L_{Cor}$ \\ 
      \hline \hline
       d 		& 7 	& 10	& 7 	& 7 	& 10 	& 7 	\\ 
       t 		& 5		& 8 	& 8 	& 5 	& 8 	& 8 	\\ 
       $|C|$	& 20 	& 200 	& 603 	& 20 	& 200 	& 600 	\\ 
       $|R|$	& 4 	& 10 	& 43 	& 4 	& 10 	& 43	\\ 
       $|S|$	& 50 	& 300 	& 21945 & 50 	& 300 	& 21945	\\ 
       $|L_s|$	& 1-5 	& 1-6 	& 1-5 	& 2-6 	& 2-6 	& 2-6 	\\ 
       \begin{tiny}$\#$Maj\end{tiny}	& N/A 	& N/A 	& N/A 	& 4 	& 10 	& 20 	\\ 
       \textsc{SSS}	& $10^{182}$ & $10^{1840}$ & $10^{6695}$ & $10^{182}$ & $10^{1840}$ & $10^{6689}$ \\
      \hline 
	  \end{tabular} 
	  \end{small}
	  \end{center}

@@

        Our results can be seen on figure 1, where each data point of the graph 
        represents the average of 5 trials. To make the scale clearer, the colour 
        on the surface corresponds to the fitness, where the highest fitnesses 
        are red. The number of copies is fixed at 1, and the proportions of 
        mutations and crossovers are shown on the horizontal axes.

        \begin{figure}[h]
  		  \centering
  		  \includegraphics[width=8cm]{GAParametersData}
  	      \caption{Effectiveness of the GA's proportion parameters}
  	      \label{graph:GAParam1}
		\end{figure}

        The graph indicates that for a fixed number of evaluations, best results are 
        achieved by selecting parameters that have the majority of individuals in the next 
        generation be generated by crossovers of the most fit schedules in the previous 
        generation. In order to achieve the very highest fitness, a few randoms or mutations 
        should also be incorporated.

@@

  \subsection{Comparison Between the Different Searches}
    We tested the 4 search methods on all 6 of our problem instances. The results 
    for two of the problems, the small and large problems, are shown on figure 2 
    and 3. The medium problem exhibited characteristics very similar to the large problem.

    All methods produced nearly indistinguishable results for the uncorrelated 
    versions. This suggests that it may be possible to use the expected internal 
    correlations more effectively that the methods we have proposed do.

@@

    \begin{figure}[h]
  	  \centering
  	  \includegraphics[width=8cm]{SMALL-TYPE2_FullAverageStats}
      \caption{Comparison of the different searches on a small data set}
	\end{figure}

	\begin{figure}[h]
  	  \centering
  	  \includegraphics[width=8cm]{HUGE-TYPE2_FullAverageStats}
      \caption{Comparison of the different searches on a large data set}
	\end{figure}

	Random search provides a benchmark by which to measure the other 3 techniques. 
	As desired, they all usually perform significantly better.

	Hillclimbing works very well for the small instance, but miserably on the large 
	problem. This is because of the vastly increased number of neighbouring schedules 
	in the large problem. Hillclimbing is able to examine a much smaller portion of 
	search space and wastes its evaluations on too many points that are too similar 
	to each other. A variant that tries some fixed number of schedules in the 
	neighbourhood may perform better.

@@

	The GA is a bit of a disappointment. It works much better than random, meaning 
	that the naive crossover method used does have some value, and it scales well 
	for different sized problem instances, which is an important characteristic. 
	For the large instance, the attained fitness of around 0.4 would be the fitness 
	of a schedule for which every student and room assign 1.4 penalty points to the 
	schedule. This is reasonable, and would perhaps be very exciting if mutate search 
	wasn't so much better than GA for this problem.

\section{Conclusion}
  \subsection{Fitness Function Modifications}
    The fitness function can be modified by any monotone function to obtain identical 
    results for the 3 non-GA methods, which only need to choose the better fitness 
    and do not use the fitness values in any important way. The GA, however, uses the 
    fitness values to decide the probability of schedules being selected and modified 
    for subsequent generations. Modifying the fitness function in a variety of ways 
    (perhaps calculating the total penalty that a schedule gets instead averaging over 
    the values given by each student and room) might improve the quality of the GA 
    approach to this problem. The fitness function used here could be modified in many 
    ways, and investigating this would be very interesting.
